---
title: 'Statistical report: Implementation, Timing and Modalities of Initiation/Up-Titration of ARNI in a Swedish real world Heart Failure population with Reduced Ejection Fraction'
subtitle: 'DRAFT'
author: 'Statistician: Lina Benson'
  
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2: 
    dev: cairo_pdf
    fig_caption: yes
    fig_height: 7
    fig_width: 8
    number_sections: yes
link-citations: yes
bibliography: references.bib
nocite: '@*'
urlcolor: blue
linkcolor: black
header-includes:
   - \usepackage{draftwatermark}
   - \usepackage{subfig}
   - \usepackage{pdflscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}
---

\clearpage
\listoftables
\listoffigures
\clearpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, include = TRUE, comment = "",
  warning = FALSE, message = FALSE, fig.pos = "H",
  fig.path = "../output/figs/"
)
options(knitr.kable.NA = "")
```

```{r adjust_directory_if_needed, include=FALSE}
# Uncomment lines below if rmd file is placed in a subdirectory
knitr::opts_knit$set(root.dir = normalizePath("../"))
```

```{r load_project}
# 1. Set options in config/global.dcf
# 2. Load packages listed in config/global.dcf
# 3. Import functions and code in lib directory

ProjectTemplate::reload.project()

cacheon <- TRUE
```             

# Data handling

## Data source

SHFDB4, https://kiheartfailure.github.io/shfdb4/, v 4.0.1. 

## Inclusion/exclusion criteria

```{r, child = "./src/flowchart.Rmd"}
```

*If a patient dies immediately after the index visit he/she will not collect a drug prescribed at the index visit
and thereby be defined as a ARNi initiation user. Throughout, the index date is 
moved 14 days forward or 30 days (the latter for substudy 1, 3 groups).

First patient in: `r min(rsdata$shf_indexdtm)` and last patient in: `r max(rsdata$shf_indexdtm)`.  

The median age (IQR) is `r rsdata %>% summarise(med = fn(median(shf_age), dig = 1),
                                             q1 = fn(quantile(shf_age, probs = 0.25), dig = 1),
                                             q3 = fn(quantile(shf_age, probs = 0.75), dig = 1)) %>%
                                   mutate(out = paste0(med, " (", q1, "-", q3, ")")) %>%
                                   pull(out)` and 
`r rsdata %>% count(shf_sex) %>%
  mutate(perc = fn(n / sum(n) * 100, 1)) %>%
  filter(shf_sex == "Female") %>%
  pull(perc)`% females.    

# Statistical analysis 

## General

All analyses were performed using `r sessionInfo()$R.version$version.string` [@r]. 
The level of significance is set to 5%, two-sided. No adjustment for multiple 
comparisons were made and therefore the results should be viewed with some care.

\clearpage

## Created variables

```{r atc}
default_kable(metalm, caption = "Treatments from Prescribed Drug Register", scale_down = F)
```

```{r td}
footnote(default_kable(targetdose, caption = "Daily target doses", scale_down = F),
  general = "Substances with - are set to missing for target doses."
)
```

ARNi use is defined as early initiation (within 14 days) if not otherwise specified. 

## Missing data

Missing data was imputed with multiple imputation (n = 10) using mice [@mice]. 
Variables included in the model are indicated in 
Table \ref{tab:tab1ss12}. Cvd/1hfh was included as the Nelson-Aalen estimator. 
The same imputated dataset is used for all substudies. 

\clearpage

# Results

For substudy 1 all analyses are performed partly for 2 groups 
(early initiation of ARNI within 14 days, no initiation/initiation>14 days) and partly
for 3 groups (early initiation of ARNI within 14 days, initiation within 15-90 days, 
no initiation/initiation>90 days).

## Substudy 1 - 2 groups

### Aims

1.	To compare outcomes in patients who initiate ARNI during hospitalization (within 14 days) 
vs. those who initiate ARNi after 14 days vs. those who do not initiate; 
2.	To assess independent predictors of early initiation.

### Propensity scores and matching

A propensity score for treatment was estimated for each patient with 
logistic regression with ARNi as outcome for each of the 10 imputed datasets using the variables 
indicated in Table \ref{tab:tab1ss12}. 1:1 matching without replacement [@match] 
was thereafter performed on the average of the resulting 10 ps [@psmatch]. Matching was 
allowed if the propensity score differed by 0.01 or less. The ability of the ps 
matching to balance the baseline characteristics was assessed by 
standardized mean differences.  

1:1 has deemed the best option when taking into account the number of patients 
retained and balance between groups. Other matching ratios: `r matchingn`.

\blandscape

### Baseline characteristics

```{r, child = "./src/tab1ss12.Rmd"}
```

\elandscape

### Associations with ARNi use

Odds ratios (95% CI) were estimated using logistic regression with ARNi use as outcome. 
Variables selected are indicated in Table \ref{tab:tab1ss12}.

```{r, child = "./src/ortabss12.Rmd"}
```

\clearpage

### Outcome analysis 

The following outcomes are considered: 

- Cardiovascular mortality/first HF hospitalization (primary endpoint)
- Cardiovascular mortality
- First HF hospitalization
- Cardiovascular mortality/Total HF hospitalization
- Total HF hospitalization
- All-cause mortality

Incidence per 1000 py was calculated with 95% Poisson confidence intervals. 

Time to first events were presented with cumulative incidence curves and 
the mean cumulative function (MCF) was used to depict the recurrent events. 

Cox proportional hazards regressions were used to evaluate the association 
between ARNi and time to first event, partly crude and partly adjusted for variables indicated in 
Table \ref{tab:tab1ss12}. The variables were selected based on clinical relevance. 
The association was also evaluated in the ps matched cohort 
(so also adjusted for variables in Table \ref{tab:tab1ss12}). 
In the matched cohort the matched pairs were modelled using a frailty term. 

Recurrent events were model using a negative binomial regression 
including the log of time as an offset in the model adjusting as above. For the matched cohort a 
generalized linear mixed-effects model (GLMM) for the negative binomial family 
with matched pair as a random effect was used. 

Data were censored at 2021-12-31, death from other causes than the event or emigration from Sweden. 

Overall cohort: The median (min-max) follow-up is 
`r rsdata %>% 
filter(shf_location == "In-patient") %>% 
summarise(med = fn(median(sos_outtime_death / 365.25 * 12), dig = 1),
                                             min = fn(min(sos_outtime_death / 365.25 * 12), dig = 1),
                                             max = fn(max(sos_outtime_death / 365.25 * 12), dig = 1)) %>%
                                   mutate(out = paste0(med, " (", min, "-", max, ")")) %>%
                                   pull(out)` months for a total of 
                                   `r rsdata  %>% filter(shf_location == "In-patient") %>% 
                                   summarise(sumpy = fn(sum(sos_outtime_death) / 365.25, dig = 0)) %>%
                                   pull(sumpy)` patient-years of follow-up.

Matched cohort: The median (min-max) follow-up is 
`r rsdata  %>% 
filter(!is.na(parpop1)) %>% 
summarise(med = fn(median(sos_outtime_death / 365.25 * 12), dig = 1),
                                             min = fn(min(sos_outtime_death / 365.25 * 12), dig = 1),
                                             max = fn(max(sos_outtime_death / 365.25 * 12), dig = 1)) %>%
                                   mutate(out = paste0(med, " (", min, "-", max, ")")) %>%
                                   pull(out)` months for a total of 
                                   `r rsdata %>% 
                                    filter(!is.na(parpop1)) %>%
                                    summarise(sumpy = fn(sum(sos_outtime_death) / 365.25, dig = 0)) %>%
                                   pull(sumpy)` patient-years of follow-up.


```{r, child = "./src/kmss12.Rmd"}
```

```{r, child = "./src/outtabss12.Rmd"}
```

\clearpage

## Substudy 1 - 3 groups

### Aims

1.	To compare outcomes in patients who initiate ARNI during hospitalization (within 14 days) 
vs. those who initiate ARNi after 14 days vs. those who do not initiate; 
2.	To assess independent predictors of early initiation.

### Baseline characteristics

```{r, child = "./src/tab1ss13.Rmd"}
```

\clearpage

### Associations with ARNi use

Odds ratios (95% CI) were estimated using logistic regression with ARNi use as outcome. 
Variables selected are indicated in Table \ref{tab:tab1ss13}.

```{r, child = "./src/ortabss13.Rmd"}
```

\clearpage

### Outcome analysis 

The following outcomes are considered: 

- Cardiovascular mortality/first HF hospitalization (primary endpoint)
- Cardiovascular mortality
- First HF hospitalization
- Cardiovascular mortality/Total HF hospitalization
- Total HF hospitalization
- All-cause mortality

Incidence per 1000 py was calculated with 95% Poisson confidence intervals. 

Time to first events were presented with cumulative incidence curves and 
the mean cumulative function (MCF) was used to depict the recurrent events. 

Cox proportional hazards regressions were used to evaluate the association 
between ARNi and time to first event, partly crude and partly adjusted for variables indicated in 
Table \ref{tab:tab1ss13}. The variables were selected based on clinical relevance. 

Recurrent events were model using a negative binomial regression 
including the log of time as an offset in the model adjusting as above. 

Data were censored at 2021-12-31, death from other causes than the event or emigration from Sweden. 

Overall cohort: The median (min-max) follow-up is 
`r rsdata %>% 
filter(pop13) %>% 
summarise(med = fn(median(sos_outtime_death30 / 365.25 * 12), dig = 1),
                                             min = fn(min(sos_outtime_death30 / 365.25 * 12), dig = 1),
                                             max = fn(max(sos_outtime_death30 / 365.25 * 12), dig = 1)) %>%
                                   mutate(out = paste0(med, " (", min, "-", max, ")")) %>%
                                   pull(out)` months for a total of 
                                   `r rsdata  %>% filter(pop13) %>% 
                                   summarise(sumpy = fn(sum(sos_outtime_death30) / 365.25, dig = 0)) %>%
                                   pull(sumpy)` patient-years of follow-up.

```{r, child = "./src/outtabss13.Rmd"}
```

```{r, child = "./src/kmss13.Rmd"}
```

\clearpage

## Substudy 2

### Aims

1.	Definition of timeline and characteristics of patients (including concomitant treatments and
their doses, e.g. ACEi/ARB, SGLT2i, MRA, BB) initiated with ARNi with focus on initiatiation
during hospitalization vs outpatient care;  
2.	Evaluation of the current status of implementation of use of sacubitril/valsartan in clinical
practice (both inpatients/outpatients) 


### Baseline characteristics

```{r, child = "./src/tab1ss2.Rmd"}
```

\clearpage

### ARNi use over time

In Figure \ref{fig:overtime-1} the denominator consists of persons that at the 
middle of the year were present in SwedeHF with the specified inclusion/exclusion criteria 
(so included in SwedeHF prior to 20xx-07-01 but not dead/emigrated as of 20xx-07-01). 
The numerator consists of the number of patients of the above with at least one ARNi 
dispension during that year. Note that this is not the substudy 2 population but 
also include patients not on ARNi at inclusion.   

In Figure \ref{fig:overtime-2} the proportion of patients that, within 14 days of 
their registration in SwedeHF, collect an ARNi prescription, are presented. 

For both Figures non-ARNi users are included in the denominator.

```{r, child = "./src/overtime.Rmd"}
```

\blandscape

## Substudy 3

### Aims

1.	To compare characteristics and outcome of patients with HFrEF according to the timing of ARNi initiation (new onset HF vs history of HF < 6 months vs HF duration $\geq$ 6 months);
2.	To assess predictors of ARNi initiation across strata of HF duration 

### Baseline characteristics

```{r, child = "./src/tab1ss3.Rmd"}
```

\elandscape

### Associations with ARNi use

Odds ratios (95% CI) were estimated using logistic regression with ARNi use as outcome. 
Separate logistic regressions were estimated for HF duration strata. Variables 
selected are indicated in Table \ref{tab:tab1ss3}.

```{r, child = "./src/ortabss3.Rmd"}
```

\clearpage

### Outcome analysis 

The following outcomes are considered: 

- Cardiovascular mortality/first HF hospitalization (primary endpoint)
- Cardiovascular mortality
- First HF hospitalization
- Cardiovascular mortality/Total HF hospitalization
- Total HF hospitalization
- All-cause mortality

Incidence per 1000 py was calculated with 95% Poisson confidence intervals. 

Time to first events were presented with cumulative incidence curves and 
the mean cumulative function (MCF) was used to depict the recurrent events. 

Cox proportional hazards regressions were used to evaluate the association 
between ARNi and time to first event, partly crude and partly adjusted for variables indicated in 
Table \ref{tab:tab1ss3}. The variables were selected based on clinical relevance. 

Recurrent events were model using a negative binomial regression 
including the log of time as an offset in the model adjusting as above. 

An interaction effect between HF duration and ARNi use was included in the model 
thereby rendering the interpretation of subgroups but with the additional advantage of also testing for 
differences in the association between ARNi and outcomes for different levels of duration of HF. 

Data were censored at 2021-12-31, death from other causes than the event or emigration from Sweden. 

The median (min-max) follow-up is 
`r rsdata %>% summarise(med = fn(median(sos_outtime_death / 365.25 * 12), dig = 1),
                                             min = fn(min(sos_outtime_death / 365.25 * 12), dig = 1),
                                             max = fn(max(sos_outtime_death / 365.25 * 12), dig = 1)) %>%
                                   mutate(out = paste0(med, " (", min, "-", max, ")")) %>%
                                   pull(out)` months for a total of 
                                   `r rsdata %>% summarise(sumpy = fn(sum(sos_outtime_death) / 365.25, dig = 0)) %>%
                                   pull(sumpy)` patient-years of follow-up.

```{r, child = "./src/kmss3.Rmd"}
```

```{r, child = "./src/outtabss3.Rmd"}
```

\clearpage

# Reproducibility

## R session information {#sessioninfo}

```{r sessinfo}
sessionInfo()
```

## R code

The R code for all data handling and statistical analyses are found: 
https://github.com/KIHeartFailure/swedehfarni. On publication
the repository will be made public so as to 
link to it from the resulting article for increased transparency and code sharing.
No data or output is stored in the repository. 

# References
